---
title: "ALDEx3 Mixed-Effects Engines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ALDEx3 Mixed-Effects Engines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Motivation

ALDEx3 fits a regression model to every Monte Carlo draw of the latent
log-abundance for every feature. For mixed-effects models, repeating a full
exact variance-component optimisation for each draw can dominate runtime.

The exact ALDEx3 mixed-effects engines remain available through
`method = "lme4"` and `method = "nlme"`. The `method = "blmm"` engine is an
ALDEx3-specific approximation designed to reduce repeated optimisation cost
while preserving the same fixed-effects model.

## What BLMM Approximates

For one feature and one Monte Carlo draw, the target remains the usual Gaussian
linear mixed model fitted in profiled REML form. BLMM does **not** approximate
the fixed-effects solve once covariance parameters are chosen.

The approximation is only this:

1. Fit one feature-specific batched anchor covariance model across all draws.
2. Build draw-specific local score updates around that anchor.
3. Update the covariance parameters per draw with a local Newton step.
4. Solve the fixed effects exactly for each draw conditional on the updated
   covariance parameters.

If the approximate path cannot be evaluated cleanly for a feature, ALDEx3 warns
and falls back to the exact `lme4` engine for that feature.

## Conceptual Formulation

Conceptually, each draw has its own REML objective for the same design matrices
and grouping structure but a different response vector. BLMM introduces:

1. One anchor fit per feature.
2. One shared local curvature matrix at the anchor.
3. Draw-specific local score corrections around the anchor.
4. Exact conditional GLS fixed-effect solves for each draw.

This is the cleanest way to describe the method statistically.

## Implemented Profiled Formulation

For one feature and one Monte Carlo draw, BLMM targets the same Gaussian linear
mixed model as `lme4`:

\[
y_s = X \beta_s + Z u_s + \varepsilon_s,
\qquad
u_s \sim N(0, \sigma_s^2 \Sigma_{\theta_s}),
\qquad
\varepsilon_s \sim N(0, \sigma_s^2 I_n).
\]

Here `X` is the fixed-effects design matrix, `Z` is the random-effects design
matrix, `\beta_s` is the draw-specific fixed-effects vector, and `\theta_s`
denotes the top-level covariance parameters. As in `lme4`, BLMM uses a
relative covariance parameterisation in which `\Sigma_{\theta_s}` is encoded
through a lower-triangular factor `\Lambda_{\theta_s}`.

The code uses the profiled `lme4`-style Gaussian LMM parameterisation instead
of a dense covariance-matrix implementation. This keeps BLMM compatible with
`lme4::lFormula()` and the relative covariance parameterisation already used by
the exact engine.

After profiling out the fixed effects and residual scale, the draw-specific
objective depends only on the covariance parameters. In the profiled
parameterisation used here, the key matrices are

\[
A_{\theta_s} = I_q + \Lambda_{\theta_s} Z^T Z \Lambda_{\theta_s}^T,
\]

\[
R_{ZX,\theta_s} = L_{\theta_s}^{-1} \Lambda_{\theta_s} Z^T X,
\qquad
M_{\theta_s} = X^T X - R_{ZX,\theta_s}^T R_{ZX,\theta_s},
\]

where `L_{\theta_s}` is the Cholesky factor of `A_{\theta_s}`. The profiled
weighted residual sum of squares for draw `s` is

\[
\mathrm{PWRSS}_s(\theta_s)
=
\|y_s\|^2
- \|R_{ZY,\theta_s}\|^2
- \|C_{\beta,\theta_s}\|^2,
\]

with

\[
R_{ZY,\theta_s} = L_{\theta_s}^{-1} \Lambda_{\theta_s} Z^T y_s,
\qquad
C_{\beta,\theta_s} = L_{X,\theta_s}^{-1}
\left(X^T y_s - R_{ZX,\theta_s}^T R_{ZY,\theta_s}\right),
\]

and `L_{X,\theta_s}` the Cholesky factor of `M_{\theta_s}`. Up to constants,
the profiled REML criterion for one draw is

\[
\ell_s(\theta_s)
\propto
\frac{1}{2}
\left\{
\log |A_{\theta_s}|
+ \log |M_{\theta_s}|
+ (n - p)\log\left(\frac{\mathrm{PWRSS}_s(\theta_s)}{n - p}\right)
\right\}.
\]

For a feature with response block `Y_d` of dimension `n x S`, one batched
objective evaluation reuses shared structure:

1. one factorisation of the profiled random-effects system,
2. one log-determinant calculation,
3. one solve against the fixed-effects design matrix,
4. one multi-right-hand-side solve against `Y_d`,
5. one average profiled residual sum-of-squares term across draws.

That batching step is the core computational advantage. BLMM does **not** loop
over `S` separate REML optimisations inside the anchor optimiser.

More concretely, the anchor fit replaces the single-draw profiled residual term
with the across-draw average

\[
\bar{\mathrm{PWRSS}}(\theta)
= \frac{1}{S}\sum_{s=1}^S \mathrm{PWRSS}_s(\theta),
\]

and optimises the batched profiled objective

\[
\bar{\ell}(\phi)
\propto
\frac{1}{2}
\left\{
\log |A_{\phi}|
+ \log |M_{\phi}|
+ (n - p)\log\left(\frac{\bar{\mathrm{PWRSS}}(\phi)}{n - p}\right)
\right\},
\]

where `\phi` is the unconstrained optimisation parameter and positive
components of `\theta` are represented on the log scale.

BLMM then uses a local second-order approximation around the anchor. Let `H` be
the observed Hessian of `\bar{\ell}` at the anchor `\bar{\phi}`. For draw `s`,
BLMM computes a draw-specific score correction `g_s` and applies the local
Newton update

\[
\phi_s \approx \bar{\phi} + H^{-1} g_s.
\]

This yields one updated covariance parameter vector per draw without rerunning
full REML optimisation from scratch. Conditional on `\phi_s`, BLMM then solves
the fixed effects exactly by generalized least squares:

\[
\hat{\beta}_s =
\left(X^T V_s^{-1} X\right)^{-1} X^T V_s^{-1} y_s,
\]

with `V_s = Z G_s Z^T + \sigma_s^2 I_n` implied by the updated covariance
parameters. The approximation is therefore confined to the repeated covariance
optimisation step; the per-draw fixed-effects solve remains exact conditional
on the updated covariance model.

## Assumptions and Scope

BLMM assumes the same Gaussian linear mixed model structure as the exact
`lme4` engine for a given feature and draw. It reuses `lme4::lFormula()` for
formula parsing and random-effects design construction, so random intercepts,
random slopes, and correlated random effects are supported through the same
model syntax.

The method is intended for ALDEx3 repeated mixed-model fitting across Monte
Carlo draws. It should be validated against exact `lme4` fits on the target
problem class before large-scale use.

## Validation Strategy

The BLMM test suite focuses on algorithmic correctness first.

1. Single-draw consistency checks compare `blmm` to exact `lme4`.
2. Identical-draw checks verify that identical Monte Carlo draws produce
   identical BLMM outputs.
3. Simulated-data comparisons check posterior means and posterior uncertainty
   against the exact engine.
4. Random-slope and correlated-random-effect tests verify that covariance terms
   are retained in the returned random-effects output.
5. Failure-mode tests verify that BLMM warns and falls back to exact `lme4`
   instead of silently changing the model.

The practical target is that approximation error in posterior means should stay
materially smaller than posterior uncertainty used for inference, and that
posterior uncertainty itself should remain close enough to avoid changing
routine interpretation.

## Worked Example

The `oral_mouthwash_data` dataset provides a repeated-measures design with one
random intercept per participant. The exact and approximate engines use the
same mixed-effects formula:

```{r eval = FALSE}
set.seed(42)
data(oral_mouthwash_data)

Y <- oral_mouthwash_data$counts
keep_names <- row.names(Y[((rowSums(Y == 0)) / ncol(Y)) <= 0.75, ])
other <- colSums(Y[((rowSums(Y == 0)) / ncol(Y)) > 0.75, ])
Y <- rbind(Y[keep_names, ], other = other)

meta <- oral_mouthwash_data$metadata

fit_blmm <- aldex(
  Y,
  ~ treat * timec + (1 | participant_id),
  data = meta,
  method = "blmm",
  nsample = 250,
  scale = clr.sm,
  gamma = 0
)

head(summary(fit_blmm))
```

Switching between exact and approximate fitting only changes the engine:
replace `method = "blmm"` with `method = "lme4"` for exact variance-component
optimisation.

## Example Comparison: `lme4` Versus `blmm`

On a small subset of the `oral_mouthwash_data` features, `blmm` tracks the
exact `lme4` results closely for most coefficients while running the
covariance-optimisation step approximately. The code below fits the same model
with both engines and compares the posterior mean coefficients and their
posterior standard errors.

```{r eval = FALSE}
set.seed(42)
data(oral_mouthwash_data)

Y0 <- oral_mouthwash_data$counts
keep_names <- row.names(Y0[((rowSums(Y0 == 0)) / ncol(Y0)) <= 0.75, ])
other <- colSums(Y0[((rowSums(Y0 == 0)) / ncol(Y0)) > 0.75, ])
Y <- rbind(Y0[keep_names, ], other = other)

# Keep a small subset to make the comparison easy to inspect in the vignette.
Y <- Y[c(seq_len(min(8, nrow(Y) - 1)), nrow(Y)), ]
meta <- oral_mouthwash_data$metadata

fit_lme4 <- aldex(
  Y,
  ~ treat * timec + (1 | participant_id),
  data = meta,
  method = "lme4",
  n.cores = 1,
  nsample = 16,
  scale = clr.sm,
  gamma = 0
)

fit_blmm <- aldex(
  Y,
  ~ treat * timec + (1 | participant_id),
  data = meta,
  method = "blmm",
  n.cores = 1,
  nsample = 16,
  scale = clr.sm,
  gamma = 0
)

s_lme4 <- summary(fit_lme4)
s_blmm <- summary(fit_blmm)

cmp <- merge(
  s_lme4,
  s_blmm,
  by = c("parameter", "entity"),
  suffixes = c(".lme4", ".blmm")
)

cmp$abs_est_diff <- abs(cmp$estimate.lme4 - cmp$estimate.blmm)
cmp$abs_se_diff <- abs(cmp$std.error.lme4 - cmp$std.error.blmm)

cmp[order(cmp$abs_est_diff, decreasing = TRUE), ][1:8, c(
  "parameter", "entity",
  "estimate.lme4", "estimate.blmm", "abs_est_diff",
  "std.error.lme4", "std.error.blmm", "abs_se_diff"
)]
```

In one run of that comparison, the median absolute difference between `lme4`
and `blmm` posterior mean coefficients was about `0.158`, and the median
absolute difference in posterior standard errors was about `0.037`. The
largest coefficient differences were concentrated in a few interaction terms
for `g__Lachnospiraceae_[G-3]` and the aggregated `other` feature, while the
standard-error differences stayed much smaller.

One representative slice from that run looked like this:

```text
                                   parameter                   entity estimate.lme4 estimate.blmm abs_est_diff std.error.lme4 std.error.blmm abs_se_diff
 treatantiseptic-mouthwash:timec15min-later g__Lachnospiraceae_[G-3]    -0.3240249    -1.5572752    1.2332503       2.260624       2.302767  0.04214379
                   treatsoda:timec2hrs-later                    other    -1.0886355    -1.8788750    0.7902395       1.423436       1.502469  0.07903349
                            timec15min-later g__Lachnospiraceae_[G-3]     0.5793505     1.3104411    0.7310906       1.598502       1.628302  0.02980016
                  treatalcoholfree-mouthwash g__Lachnospiraceae_[G-3]    -1.6565051    -1.0162586    0.6402465       1.836860       1.904320  0.06746007
                                   treatsoda                    other     0.6267303     1.2657432    0.6390129       1.378232       1.422042  0.04381026
 treatalcoholfree-mouthwash:timec15min-later g__Lachnospiraceae_[G-3]     0.7208208     0.0866997    0.6341211       2.390340       2.436982  0.04664163
                  treatsoda:timec15min-later                    other    -0.4414167    -1.0440975    0.6026808       1.385616       1.462925  0.07730838
   treatantiseptic-mouthwash:timec2hrs-later g__Lachnospiraceae_[G-3]     0.2297808    -0.3268711    0.5566519       2.260624       2.302767  0.04214379
```

That pattern is the practical target for BLMM: most coefficient and standard
error summaries should stay close to the exact engine, with approximation error
small relative to the posterior uncertainty being propagated through ALDEx3.
